\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-05 sections/chameleon.tex
John's comments
A temporary save

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/single-error-location-example.png}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
    }
    \label{fig:single-candidate}
\end{figure}

>>>>

\todo{I wonder about a high level overview diagram of tool / usage of tool, then the details of each component?}

\todo{Can you use this example in Motivating Example sectoon (see my comment above) then use it to explain tool in this section?  Make sense?

Is this example too simple...?}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/add-pair-example.pdf}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
% \todo{this figure unreadable...}
    }
    \label{fig:single-candidate}
\end{figure}

### 2022-12-05 sections/evaluation.tex
A temporary save

>>>>

\subsection{Threats to validity}

% \todo{Put threats BEFORE Discussion and at end Eval section?}

\todo{These are too terse/unclear - need expanding. Suggest classify as internal / external threats etc .  What about background and number of participants as threats??}

One consequence of our recruiting approach is it is harder for previous participants to enter a later study. To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.

Conducting studies remotely and unsupervised is that intervening when users encounter usability issues mid-study is impossible. To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

### 2022-12-05 sections/introduction.tex
update from overleaf
A temporary save

Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

Programmers find challenges in adopting typed languages in their practice. They tend to be verbose, hard to learn and generate bad error messages. 
One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues pose challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.  In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions. The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

@@

This paper makes the following contributions:
\begin{itemize}
\item { We provide the design and implementation of the \chameleon{}, to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  }
\item {
    We report the result of four experiments designed to evaluate \chameleon{}.}
\end{itemize}

>>>>

% \todo{As above, would start with motivation of dynamically typed langs e.g. JavaScript, Pythjon popularity but issues this brings...
% }

Dynamically typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}. These languages have the advantage of enormous flexibility: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection. However, studies show that dynamically typed languages show challenges in development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}. They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and reuqire strong discipline to avoid pitfalls \cite{chen_empirical_2020}. For these reasons, many modern dynamically typed languages have introduced statically typing as part of the core language features in recent years.

Functional programming languages have long enjoyed rigorous type systems and expressive type-level features. Techniques such as type inference and algebraic types have long been standard practice in functional languages such as ML and Haskell, and now fresh blood in imperative languages, such as Rust and typescript. Haskell is considered a "type-system laboratory" \cite{hudak_history_2007}. Various type system advancements were experimented with in Haskell and ended up in mainstream languages years or even decades after. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

The downside of programming in functional, statically typed language is its complexity, especially when encountering type errors. Understanding and fixing type errors are shown to be hard\cite{tirronen_understanding_2015}. In particular, they may point to locations that are distant from real error causes, expose errors in hard to comprehend, or provide misleading fixing suggestions. In addition, the tools equipped are generally ineffective in showing and explaining type errors\cite{wu_how_2017}.
% Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

% \todo{There are a lot of minor grammer fix-ups needed throughout...}
% Programmers find challenges in adopting typed languages in their practice. \todo{I think this statement is plain wrong myself... 
%  Need strong references to justify if going to say this to SE audience!:}They tend to be verbose, hard to learn and generate bad error messages. 
% One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues pose challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

This paper introduces \chameleon{}, an interactive debugging tool for Haskell. It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it). In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature. These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

@@

This paper makes the following contributions:
\begin{itemize}
\item We provide the design and implementation of the \chameleon{} to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  
\item {
    We report the result of four experiments designed to evaluate \chameleon{}.}
\end{itemize}
